---
title: "ch_5__data_transformation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nycflights13)
library(tidyverse)
```


exploring the basic dataframe

```{r}
flights
```

Notes from the book: 

The verbs to know: 

1. Pick observations by their values (filter()).
2. Reorder the rows (arrange()).
3. Pick variables by their names (select()).
4. Create new variables with functions of existing variables (mutate()).
5. Collapse many values down to a single summary (summarise()).

All verbs work similarly:

1. The first argument is a data frame.
2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).
3. The result is a new data frame.

Together, the verbs make it easy to chain together multiple simple steps to achieve a complex result. 

##5.2 filtering
```{r}
#filter the flights data for the first month and first day
filter(flights, month == 1, day == 1)
```

wrapping a command in `()` will run the results AND show the output. Note the difference:

```{r}
dec25 <- filter(flights, month == 12, day == 25)

(dec25 <- filter(flights, month == 12, day == 25))
```

using `OR` aka `|`

```{r}
#this doesn't work to find flights in november OR december
filter(flights, month == 11 | 12)
```

here's what DOES work:
```{r}
filter(flights, month == 11 | month == 12)
```

and alternatively: 
```{r}
filter(flights, month %in% c(11, 12))
#this is the first time we see %in%
#select every row where [x] is one of the values in the c() grouping. 
```

and because that works and it looks like the book wants to use it later:
```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```

###missing values

```{r}
# Let x be Mary's age. We don't know how old she is.
x <- NA

# Let y be John's age. We don't know how old he is.
y <- NA

# Are John and Mary the same age?
x == y
#> [1] NA
# We don't know!
```

If you want to determine if a value is missing, use is.na():
```{r}
is.na(x)
```

*filter() only includes rows where the condition is TRUE*

###5.2.4 exercises:
Find all flights that:

1. Had an arrival delay of two or more hours
```{r}
#we need the flights data
flights %>% 
  #and arrival delay is in minutes, so filter that to greater than 120 minutes
  filter(arr_delay >= 120)
```

2. Flew to Houston (IAH or HOU)
```{r}
flights %>% 
  filter(dest %in% c("IAH", "HOU")) %>% 
  arrange(carrier)
```

3. Were operated by United, American, or Delta
```{r}
flights %>% 
  filter(carrier %in% c("UA", "AA", "DL"))
```

4. Departed in summer (July, August, and September)
```{r}
flights %>% 
  filter(month %in% c(7,8,9))
```

5. Arrived more than two hours late, but didn’t leave late
```{r}
flights %>% 
  filter(arr_delay > 120,
         dep_delay <= 1)
```

6. Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
flights %>% 
  filter(dep_delay >= 60,
        (dep_delay - arr_delay) >=30)
```

7. Departed between midnight and 6am (inclusive) 
```{r}
flights %>% 
  filter(dep_time >= 0,
         dep_time <= 600) %>% 
  arrange(desc(dep_time))
```

8. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
```{r}
flights %>% 
  filter(between(dep_time, 0, 600)) %>% 
  arrange(desc(dep_time))
```
9. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
```{r}
flights %>% 
  filter(is.na(dep_time))
```

Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)


##5.3 Arranging

arrange changes orders. `arrange()` can be nested to arrange by this, then that, then another thing. any additional arranged value beyond the first is used to break ties. Missing values are always sorted at the end!

###5.3.1 Exercises
How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
flights %>% 
  arrange(desc(is.na(dep_time)))
```

Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
flights %>% 
  arrange(desc(dep_delay))
```

Sort flights to find the fastest flights.
```{r}
flights %>% 
  arrange(arr_delay-dep_delay)
```

Which flights travelled the longest? Which travelled the shortest?
```{r}
flights %>% 
  arrange(distance)

flights %>% 
  arrange(desc(distance))

```

##5.4 Select columns with select()

```{r}
flights %>%
  rename(tail_num = tailnum)
```

###Exercises 
Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

What happens if you include the name of a variable multiple times in a select() call?
```{r}
flights %>% 
  select(tailnum, tailnum)
```

What does the one_of() function do? Why might it be helpful in conjunction with this vector?

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

flights %>% 
  select(one_of(vars))
```
Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
```{r}
select(flights, contains("TIME"))
```

##5.5 Add new variables with `mutate()`
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time) 
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```

###5.5.2 Exercises
Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
flights %>% 
  select(dep_time, sched_dep_time) %>% 
  mutate(dep_hr_mins = (dep_time %/% 100)*60,
         dep_minute = dep_time %% 100,
         sched_hr_mins = (sched_dep_time %/% 100)*60,
         sched_mins = sched_dep_time %% 100) %>% 
  transmute(dep_mins_since_midnight = dep_minute + dep_hr_mins,
            sched_dep_mins = sched_hr_mins + sched_mins)
```

Compare `air_time` with `arr_time` - `dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?
```{r}
flights %>% 
  select(air_time, arr_time, dep_time) %>% 
  mutate(flight_time = arr_time - dep_time)
#The problem is continuous numbers. dep_time and arr_time are numerical hmm; and would need to be converted like the previous problem. 
```

Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?
```{r}
flights %>% 
  select(dep_time, sched_dep_time, dep_delay)
#I expect that dep_delay is the number of minutes that a flight was delayed. negative numbers mean the dep_time happened BEFORE the sched_dep_time
```

Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().
```{r}
flights %>% 
  mutate(worst_delay = min_rank(desc(arr_delay))) %>% 
  arrange(worst_delay) %>% 
  filter(worst_delay <= 10)
#ties can be broken by arranging using two factors.
```

What does 1:3 + 1:10 return? Why?
```{r}
(1:3 + 1:10)
#returns "longer object length is not a multiple of shorter object length [1]  2  4  6  5  7  9  8 10 12 11" because the first object length is shorter and loops through when it reaches it's end.
```

What trigonometric functions does R provide?
```{r}
#base::Trig is the fucntion set specific to trig. It contains
#cos(x)
#sin(x)
#tan(x)

#acos(x)
#asin(x)
#atan(x)
#atan2(y, x)

#cospi(x)
#sinpi(x)
#tanpi(x)
```

##5.6 Grouped summaries with summarise()
Summarise collapses data into a single row; but really only works when paired with `group_by()`.  

in the book example: 
```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```
this groups according to each of those factors, then summarises according to the smallest group that we grouped_by. 

###5.6.1 Combining multiple operations with the pipe
heh.

So, before the pipe, we had to name each of the smaller data frames as we build toward our final data frame; like this: 
```{r}
#first we group
by_dest <- group_by(flights, dest)

#then we summarise(ze)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE))

#then we filter
delay <- filter(delay, count > 20, dest != "HNL")

#then we visualize
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE) +
  theme_light() 
```

alternatively, we can use the pipe (it's not a pipe) to do all of this in "one" step.

the pipe is one of the key criteria for belonging to the tidyverse.
```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

###5.6.2 Missing values
what happens if we dont use `na.rm`? Look at all these `NA`'s!
```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

NOTE: aggregation functions obey the usual rule of missing values: if there’s any missing value in the input, the output will be a missing value. so we use `na.rm` to get rid of the `NA` values.

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))
```

```{r}
#save the "is not NA" dataset
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

###5.6.3 Counts
Always good to do counts whenever summarising; ANY aggregation; to make sure you aren't using small samples. 

    * "it’s always a good idea to include either a count (n()), or a count of non-missing values (sum(!is.na(x)))"

When looking at plots of summaries vs counts (group size), variation increases as sample size decreases. Check this, and if you see this happening, filter out groups with small observations. This shows more of the pattern and less of the small-sampling-variation. 

```{r}
#using the delays dataset
delays %>% 
  #filter out any values with a count less than 25
  filter(n > 25) %>% 
  #then show me a scatterplot
  ggplot(mapping = aes(x = n, y = delay)) + 
    geom_point(alpha = 1/10)
```

Here's another example from the book: 

```{r}
#install.packages("Lahman")
#load Lahman
library(Lahman)

# Convert to a tibble so it prints nicely
batting <- as_tibble(Lahman::Batting)

batters <- batting %>% 
  #group all at bats by the playerID
  group_by(playerID) %>% 
  #summarise the batting average and the at bats
  #remembering to remove rows with NA values
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>% 
  filter(ab > 30) %>% 
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() + 
    geom_smooth(se = FALSE)
#> `geom_smooth()` using method = 'gam' and formula 'y ~ s(x, bs = "cs")'
```

(You can find a good explanation of this problem at http://varianceexplained.org/r/empirical_bayes_baseball/ and http://www.evanmiller.org/how-not-to-sort-by-average-rating.html.) 

###5.6.4 Useful summary functions

Also useful:

    * median(x)
    * measures of spread: sd(x), IQR(x), mad(x): _Standard Deviation_ is a common measurement of spread. Sometimes the interquartile range or median absolute deviation is better if we have outliers. 
    * Measures of rank: min(x), quantile(x, 0.25), max(x): for quantile, the decimal is the percentage.
    * Measures of position: first(x), nth(x, 2), last(x)
    * Counts: n(), which takes no arguments, and returns the size of the current group. To count the number of non-missing values, use sum(!is.na(x)). To count the number of distinct (unique) values, use n_distinct(x). Or even count() -- this is a dplyr verb

###5.6.5 Grouping by multiple variable

"When you group by multiple variables, each summary peels off one level of the grouping. That makes it easy to progressively roll up a dataset"

```{r}
# using the flights dataset build a dataset that's first grouped by year, then month, then day
daily <- group_by(flights, year, month, day)
# count the number of flights per day, this automatically includes month and year
(per_day   <- summarise(daily, flights = n()))
```

```{r}
# use the same data and aggregate by month, which automatically rolls the year grouping as well
(per_month <- summarise(per_day, flights = sum(flights)))
```

```{r}
(per_year  <- summarise(per_month, flights = sum(flights)))
```

it's cool to roll up summaries for counts and sums, but this is NOT ok with rank-based stats like medians. Give a hard think about means and variances. "the sum of groupwise sums is the overall sum, but the median of groupwise medians is not the overall median." so says the book.

###5.6.6 Ungrouping

ungroup() does what it says, ungroups grouped data. 

###5.6.7 Exercises
