---
title: "ch_5__data_transformation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nycflights13)
library(tidyverse)
```

exploring the basic dataframe

```{r}
flights
```

Notes from the book: 

The verbs to know: 

1. Pick observations by their values (filter()).
2. Reorder the rows (arrange()).
3. Pick variables by their names (select()).
4. Create new variables with functions of existing variables (mutate()).
5. Collapse many values down to a single summary (summarise()).

All verbs work similarly:

1. The first argument is a data frame.
2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).
3. The result is a new data frame.

Together, the verbs make it easy to chain together multiple simple steps to achieve a complex result. 

##filter
```{r}
#filter the flights data for the first month and first day
filter(flights, month == 1, day == 1)
```

wrapping a command in `()` will run the results AND show the output. Note the difference:

```{r}
dec25 <- filter(flights, month == 12, day == 25)

(dec25 <- filter(flights, month == 12, day == 25))
```

using `OR` aka `|`

```{r}
#this doesn't work to find flights in november OR december
filter(flights, month == 11 | 12)
```

here's what DOES work:
```{r}
filter(flights, month == 11 | month == 12)
```

and alternatively: 
```{r}
filter(flights, month %in% c(11, 12))
#this is the first time we see %in%
#select every row where [x] is one of the values in the c() grouping. 
```

and because that works and it looks like the book wants to use it later:
```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```

##missing values

```{r}
# Let x be Mary's age. We don't know how old she is.
x <- NA

# Let y be John's age. We don't know how old he is.
y <- NA

# Are John and Mary the same age?
x == y
#> [1] NA
# We don't know!
```

If you want to determine if a value is missing, use is.na():
```{r}
is.na(x)
```

*filter() only includes rows where the condition is TRUE*

##exercises:
Find all flights that:

1. Had an arrival delay of two or more hours
```{r}
#we need the flights data
flights %>% 
  #and arrival delay is in minutes, so filter that to greater than 120 minutes
  filter(arr_delay >= 120) 
```

2. Flew to Houston (IAH or HOU)
```{r}
flights %>% 
  filter(dest %in% c("IAH", "HOU"))
```

3. Were operated by United, American, or Delta

4. Departed in summer (July, August, and September)
5. Arrived more than two hours late, but didn’t leave late
6. Were delayed by at least an hour, but made up over 30 minutes in flight
7. Departed between midnight and 6am (inclusive)
8. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
9. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)

```{r}
```

